---
title: "Factors"
author: "Sean Davis & Lori Kern"
---

## Factors

A factor is a special type of vector, normally used to hold a
categorical variable--such as smoker/nonsmoker, state of residency, zipcode--in many statistical functions. Such vectors have class “factor”. Factors are primarily used in Analysis of Variance (ANOVA) or other situations when "categories" are needed. When a factor is used as a predictor variable, the corresponding indicator variables are created (more later).

Note of caution that factors in R often *appear* to be character vectors
when printed, but you will notice that they do not have double quotes
around them. They are stored in R as numbers with a key name, so
sometimes you will note that the factor *behaves* like a numeric vector.

```{r}
# create the character vector
citizen<-c("uk","us","no","au","uk","us","us","no","au") 

# convert to factor
citizenf<-factor(citizen)                                
citizen             
citizenf
# convert factor back to character vector
as.character(citizenf)
# convert to numeric vector
as.numeric(citizenf)
```

R stores many data structures as vectors with "attributes" and "class" (just so you have seen this).

```{r}
attributes(citizenf)
class(citizenf)
# note that after unclassing, we can see the 
# underlying numeric structure again
unclass(citizenf)
```

Tabulating factors is a useful way to get a sense of the "sample" set available. 

```{r}
table(citizenf)
```

The default factor levels are the unique set of possible values. It is possible
to specify a subset of factor levels. Note how missing values are introduced if
a value is not included.

```{r}
citizenf2 <- factor(citizen, levels=c("us", "uk"))
citizenf2

table(citizenf2)
```

Missing values are exlcuded by default. There is an option to override this
setting.

```{r}
addNA(citizenf2)
table(addNA(citizenf2))
```

::: {.callout-caution}
This emphasizes that default settings may or may not be appropriate for your
analysis. It's important to know what those settings are and choose alternatives
as necessary.
:::
