---
author: "Sean Davis and Lori Kern"
---
# Saving and Loading Workspaces and Objects

## Rstudio Projects

**placekeeper**


## Saving and Loading Workspace

If you are not in Rstudio and want to save your workspace there are a few
options. If you choose to save your session when you quit out of R
with `q("yes")` or `q()` and selecting `yes`, it is also equivalent to the
following using save

```{r, eval=FALSE}
save(list = ls(all.names = TRUE), file =".RData", envir = .GlobalEnv)
```
This saves your workspace and command history. Any object assignments and
command history (viewable with `history()`) will be available in your next R
session. Your session will load automatically when you start up R in the same
directory. 

You can use the `save.image` option

```{r, eval=FALSE}
save.image()
# or
save.image(file="descriptiveFileName.RData")
```

If you do not give `save.image` a file name, it will also load on default if you
start an R session in that directory but with default settings will not have
your command history. 

If you give `save.image` a file name, it will not automatically load when you
start R and will not have command history. You will have to load the image
manually to see the objects in the new R session.

```{r, eval=FALSE}
load("descriptiveFileName.RData")
```

## Saving and Loading Objects

If you want to selectively save objects, the `save` function is utilized. This
allows you to choose the objects you want saved. The `load` function would then
load the selected objects into a new R session.

```{r, eval=FALSE}
ages = 1:4
months = c("may", "june", "july", "august")
vec = c(TRUE, FALSE, TRUE)
save(months, ages, file="subset.RData")
```

## Saving and Loading Command History

When quitting out of R, if you save, it also saves your command history. To do
this manually you can use the `savehistory`/`loadhistory` functions in R.   